---
title: "BIO456/BIO556 Functions & Loops & Things"
date: "11 Sept 2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

In this RMarkdown, we practice using FUNCTIONS, LOOPS and a smattering of other things...

Overview
I. Last week's lab assignment
II. R tidbits (random useful things)
III. If/else statements
IV. Intro functions
V. Intro for() loops
VI. Combining for() loops and if() statements
VII. While loops
VIII. Example exercise, re: lecture
IX. LAB EXERCISE
X. Bonus content

As always, let's load in our packages:
```{r libraries, message=FALSE}
library(tidyverse)
```

## I. First let's go over the answer to the lab assignment from last week... 

Things we wanted to do:

1 - Filter the sticky trap data to include only Diptera observations on traps that were out for less than six days

2 - Tidy the ID effort data, i.e. one column for observer names and another for presence/absence

3 - Summarize the number of observers per traps from each combination of date and treatment

4 - Combine the sticky trap and ID effort data and feed the result into the ggplot code I shared

```{r lab 02 review STEP 1}
#STEP 1: Filter the sticky trap data
sticky_traps <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/Labs/Lab2/sticky_traps.csv") %>%
  select(trap_ID, days_out, insect_order, count) %>%   # optional cleaning (removes extra columns)
  filter(insect_order == "Diptera" & days_out < 6) # non-optional filtering, as specified above
```

```{r lab 02 review STEP 2}
#STEP 2: Tidy the ID effort data
ID_effort <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/Labs/Lab2/ID_effort.csv") #read in data

ID_effort <- ID_effort %>% pivot_longer(Avalon:Belle, names_to = "observer", values_to = "presence") # tidy data, wide to long
```

```{r lab 02 review STEP 3}
#STEP 3: Summarize the total effort per trap using code I shared
ID_effort <- ID_effort %>% group_by(date, treatment) %>% summarize(effort = sum(presence))      
```

```{r lab 02 review STEP 4}
#STEP 4: Combine the data on a common column
#First, make sure there is a common column on which to combine the data
sticky_traps
ID_effort

sticky_traps <- sticky_traps %>% separate(trap_ID, into = c("date","treatment"))      

#Combine the two tibbles
sticky_combo <- sticky_traps %>% left_join(ID_effort)    

#Run provided ggplot code
ggplot(sticky_combo, aes(x = treatment,                  # light treatment (dark, red, amber, or blue)
                         y = count/days_out,             # number of Diptera caught / number of days trap was out
                         fill = effort %>% as.factor)) + # effort will be a number, but we want it as a category
  geom_jitter(height = 0, width = 0.25, shape = 21, size = 2, alpha = 0.75) + 
  scale_fill_manual(name = "Number of observers", 
                    values = c("darkred", "pink", "dodgerblue", "greenyellow")) +
  labs(x = "Light treatment", y = "Attraction to sticky traps (number caught/nights out)") +
  theme_bw()
```

Review of functions in our code above:

### What do the three functions, group_by(), summarize() and mutate() actually do?

```{r summarize()}
leafhoppers <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2024/Labs/Week1/leafhoppers.csv") %>% 
  pivot_longer(plant_01:plant_05, 
               names_to = "plant_ID", 
               values_to = "no_leafhoppers")

#Remember to always look at the data first!
leafhoppers

# use summarize() function to add all values in no_leafhoppers column together
leafhoppers %>% 
  summarize(total_bugs = sum(no_leafhoppers)) 
```

```{r group_by() & summarize() option 1}
leafhoppers %>% 
  group_by(plant_ID) %>% # use group_by() to tell R to run the code separately for each of the five different levels of the plant_ID factor
  summarize(total_bugs = sum(no_leafhoppers)) # so here we get five sums for the five different plants
```

```{r group_by() & summarize() option 2}
leafhoppers %>% 
  group_by(no_leafhoppers) %>% # what if we group by the different values in the no_leafhoppers column?
  summarize(total_bugs = sum(no_leafhoppers)) # we only have two (1 and 0), and the zeros add to zero...

# compare the above output to no group_by()
leafhoppers %>% summarize(total_bugs = sum(no_leafhoppers))
    
# compare the above output to grouping by plant_ID                    
leafhoppers <- leafhoppers %>% group_by(plant_ID)
leafhoppers %>% summarize(total_bugs = sum(no_leafhoppers))

#Be careful about what R is doing "under the hood"... If you *save* a tibble/dataframe with the data in grouped as I just did above, all future commands will be on those groups. You must UNGROUP to again run functions without application the groups.

leafhoppers <- leafhoppers %>% ungroup(plant_ID)
leafhoppers %>% summarize(total_bugs = sum(no_leafhoppers))
```

```{r mutate()}
sticky_traps <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/Labs/Lab2/sticky_traps.csv") %>% 
  select(trap_ID, days_out, insect_order, count) %>% 
  separate(trap_ID, into = c("date", "treatment")) 

#again, look at the data!
sticky_traps

sticky_traps %>% 
  mutate(observation = row_number(),     # the mutate function creates new columns
         count_per_day = count/days_out) # based on whatever you put in (name followed by formula or function)

#be sure to press the triangle at the top right of the tibble below if you cannot see the column labeled "count_per_day"

# We can also make a new column the old-fashioned way in base R. In this approach, the new column is automatically saved to our dataframe or tibble, whereas above it is not.
sticky_traps$count_per_day <- sticky_traps$count/sticky_traps$days_out
st

#check to see your new column
sticky_traps
```

```{r group_by() & mutate()}
sticky_traps %>% 
  group_by(treatment) %>%                # if you group_by() some factor or combination of factors
  mutate(observation = row_number(),
         #,     # it will run the code afresh for the different factor levels
         count_per_day = count/days_out) 
```

```{r group_by() mutate() & summarize()}
sticky_traps %>% 
  group_by(treatment, days_out, insect_order) %>% # try deleting one or more of the variables in the group_by() statement and seeing what changes!
  summarize(total_insects = sum(count)) %>%       # here, you get one sum for every combo of treatment and order
  mutate(count_per_day = total_insects/days_out) 
```
## Alternative way of plotting last week's lab

```{r Alternate plot for last weeks lab}
sticky_traps <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/Labs/Lab2/sticky_traps.csv") %>%
  select(trap_ID, days_out, insect_order, count) %>% 
  filter(insect_order == "Diptera" & days_out < 6) %>%   
  group_by(trap_ID, insect_order, days_out) %>%       # LINE OF CODE NOT IN OUR ORIGINAL APPROACH
  summarize(count = sum(count)) %>%                    # LINE OF CODE NOT IN OUR ORIGINAL APPROACH 
  separate(trap_ID, into = c("date","treatment"))

sticky_traps #I think this tibble makes a better plot -- what do you think and why?

ggplot(sticky_traps %>% left_join(ID_effort), # BTW we can actually do the join *in* the ggplot code
       aes(x = treatment, y = count/days_out, fill = effort %>% as.factor)) +  
  geom_jitter(height = 0, width = 0.25, shape = 21, size = 2, alpha = 0.75) + 
  scale_fill_manual(name = "Number of observers", 
                    values = c("darkred", "pink", "dodgerblue", "greenyellow")) +
  labs(x = "Light treatment", y = "Attraction to sticky traps (number caught/nights out)") +
  theme_bw()
```


## II. Below is a chunk full of random useful(?) things to think about

```{r just some notes}
#Create a vector
just_some_nums <- c(1337, NA, Inf, -10, 2.5)

# (1) What are each of the lines of code below doing differently?
# useful arguments: na.rm and finite
range(just_some_nums) 
range(just_some_nums, na.rm = T)
range(just_some_nums, finite = T)


# (2) What is different about the three lines of code using the sample() function below?

# randomly pick some number of elements (`size`) from the object named `x` that is a vector
sample(x = 1:6, size = 1) 

# randomly pick some number of elements (`size`) from the object named `x` that is a vector
sample(1:6, 6)

# what does replace = T do? If you're not sure, run the line of code several times
sample(1:6, size = 6, replace = T)


# (3) What is different about the two lines of code below? 

# What is different about the functions themselves? (HINT: Use the help() function or ?functionname)

# Are they doing the same thing?

str_c("the letter ", letters[1], " is the best")
paste("the letter", letters[1], "is the best")


# (4) What is R returning in the lines of code below?

letters %in% c("l","e","o","n","e","b")
letters %in% c("L","e","o","n","e","B")

LETTERS %in% c("L","e","o","n","e","B")

```

## III. If/else statements

What if you want to vary output based on the kind of input? if(), else() and ifelse()

- Decode the loops below. What is each loop doing?? How do they work?

```{r ifelse()}
# Short & efficient but kind of inflexible ways of varying output based on the kind of input:
# LOOP 1: 
ifelse(10 == 1,                                  # use `==` to test for equivalency
       yes = "criteria fulfilled",               # use `=` like `<-` for assignment
       no = "error error beep boop oh no")


# LOOP 2:
ifelse(letters %in% c("L","e","o","n","e","B"), 100, NA)
```


```{r if(){} and else{}}
# LOOP 3: Longer but more flexible if()...else() code
# First create variables that will be referred to in the loop
tiny_number <- 0.001
norm_number <- 10
huge_number <- 1000

# loop 3 version 1
if (tiny_number != huge_number) { # if tiny_number is not equal to huge_number
  print("that's good")            # then we are telling R to print this phrase to the console
  } else {                        # if tiny_number is NOT not equal to huge_number (AKA is equal to, '==')
    print("oh no...")             # then we want R to tell us this instead
    }

#TASK 1: Adjust the above if()...else() function to change the output to "oh no..."

#TASK 2: Adjust the if()...else() function below to change the output to "w- what...?!"

# loop 3 version 2
if (tiny_number == norm_number | huge_number >= norm_number) { 
  print("it's all a-OK")
  } else {
    print("w- what...?!")
    }
```


```{r nested if(){} and else{}}
### Nested if()/else() statements
tiny_number <- 2
norm_number <- 2
huge_number <- 2

if (tiny_number < norm_number & huge_number > norm_number) { # & for my final trick, nested if else statements
  print("seems legit")                                       # print this if both statements are true
  } else {                                                   # three ways not to get left == TRUE & right == TRUE
    if (tiny_number < norm_number) {                         # if the left statement is true, the right must not be
      print("problem on right hand side")                    # in which case write this
      } else { 
        if (huge_number > norm_number) {                     # if the right statement is true, the left must not be
          print("problem on left hand side")                 # in which case write this
          } else {
            print("problems all around")                     # if none of the above worked, BOTH must == FALSE
            }
      }
  }

# TASK: Adjust the values of the above named objects (tiny_number <- 0.001, norm_number <- 10, huge_number <- 1000) to entice R to print the other three options?
```

## IV. Intro functions

If you ever end up copy-pasting a lot of code over and over again, you could probably benefit from a function...

```{r function(){}}
# Example function named roll_d6(), which takes one input called no_dice
roll_d6 <- function(no_dice){              
  sample(1:6, size = no_dice, replace = T) # roll_d6(no_dice = 10) will do this: sample(1:6, size = 10, replace = T)
}                    


# Example function that will do a virtual coin flip some number of times
flip_coin <- function(times){              # this function will do a virtual coin flip some number of times
  sample(c("heads", "tails"), size = times, replace = T) # samples randomly from two options: "heads" | "tails"
}


roll_d6(3)    # What is this output giving us???
flip_coin(3) # What is this output giving us???

# Vary the inputs for the roll_d6() and flip_coin() functions to see the different outputs

#What happens if we change the code for the flip coin function to remove the "replace = T" argument?
flip_coin2 <- function(times){              
  sample(c("heads", "tails"), size = times)
}

flip_coin2(1)
flip_coin2(2)
flip_coin2(3) #Why the error??

```

If you ever end up copy-pasting a lot of code over and over again, you might also benefit from a *for loop*!

## V. Intro for() loops

```{r for(){}}
# a vector with six names in it:
DnD_crew <- c("Eriberto", "Kylie", "Helen", "Chelsea", "Ian", "Carly") 
# for() loop example 1.1:
for(x in 1:6){        # run this loop six times, with six values for x (1 through 6)
  print(DnD_crew[x])  # each time, print the contents of that place in this vector
}

# for() loop example 1.2:
for(x in 2:7){        # run this loop six times, with six values for x (2 through 7)
  print(DnD_crew[x])  # each time, print the contents of that place in this vector
}

# for() loop example 1.3:
for(x in c(1, 3, 7)){
    print(DnD_crew[x])  # each time, print the contents of that place in this vector
}

# If you can't understand what the loop is doing, break it down by assigning a value to x and just running the code within the loop to see the output:
x<-6
print(DnD_crew[x])
DnD_crew[c(1, 3, 7)]


# for() loop example 2:
# first make a vector containing the first two numbers of a Fibonacci sequence (our starting values)
fibonacci <- c(0, 1)  

# create a loop that starts with the input vector fiboacci
for(x in 2:50){       # run this loop 49 times, with 49 values for x (2 through 50)
  fibonacci <- c(fibonacci,                     # each time, add one NEW place to the old vector
                 fibonacci[x] + fibonacci[x-1]) # namely, the sum of the last two places of that vector 
}

fibonacci #output of the above loop

# Again, to understand what a loop is doing, or to troubleshoot your own loop that you might be creating,
# (1) start with setting a fixed value of "x" in the for() loop, e.g.,
x = 2

# (2) then run the code contained in the loop and see what the output is for your fixed value of x
# First, reset fibonacci to its original values:
fibonacci <- c(0, 1)  

# now, when x = 2, fibonacci =...
(fibonacci <- c(fibonacci, fibonacci[x] + fibonacci[x-1]))

# when x = 3, fibonacci = ...
x = 3
(fibonacci <- c(fibonacci, fibonacci[x] + fibonacci[x-1]))

# when x = 4, fibonacci = ...
x = 4
(fibonacci <- c(fibonacci, fibonacci[x] + fibonacci[x-1]))

# when x = x + 1, fibonacci = ...
x = x + 1
(fibonacci <- c(fibonacci, fibonacci[x] + fibonacci[x-1]))

```


Why not combine loop types?
if(), else() and for()

## VI. Combining for() loops and if() and/or else() statements

```{r functions for loops & if statements}
# functions within loops
roll_d20 <- function(x){ # here I've made a function for "rolling" a 20-sided die
  sample(1:20, size = x, replace = T)
}

roll_d20(1) #roll the 20-sided dice once
roll_d20(4) #roll the 20-sided dice four times

# Place the above function within the loop below:
for(x in 1:length(DnD_crew)){ # our DnD_crew vector has seven places, so this is the same as 'x in 1:7'
  print(str_c(DnD_crew[x], " rolled a ", roll_d20(1))) # this code has each person roll one d20
}

# What is this line of code below doing? Why is it only rolling the die once?
str_c(DnD_crew[1:6], " rolled a ", roll_d20(1))

# same output as: 
paste(DnD_crew[1:6], "rolled a", roll_d20(1))
# but why doesn't this do the same thing as the for loops above??

# But neither str_c() nor paste() do what we want them to -- they only roll the d20 once for each person!


#INTERPRET THE OUTPUTS OF THE FOLLOWING LINES OF CODE:
str_c(DnD_crew[1:8], " rolled a ", roll_d20(1)) 
str_c(DnD_crew[1:6], " rolled a ", roll_d20(6))
str_c(DnD_crew[1:6], " rolled a ", roll_d20(3)) 
str_c(DnD_crew[1:6], " rolled a ", roll_d20(7)) 

# Below we combine if() and else() statements AND a function within a loop
# We'll use this modified loop to correct our grammar print 'an' in cases where the number output by the die starts with a vowel
for(x in 1:length(DnD_crew)){                      
  roll = roll_d20(1)                               # save (& overwrite) the rolls made in all six loop iterations
  if(roll %in% c(8,11,18)){                        # check if the number rolled this iteration begins with a vowel
    print(str_c(DnD_crew[x], " rolled an ", roll)) # if so, we tell R to print 'an'
  } else {
    print(str_c(DnD_crew[x], " rolled a ", roll))  # if not, we tell R to print 'a'
  }
}

# loops within functions
# this new function is going to take TWO separate inputs
flip_coins <- function(trials, replicates){   
  outcome <- tibble(flip = 1:trials)          # this is just a placeholder; we'll overwrite this, but need something there to overwrite 
  summary <- tibble(all_tails = 1:replicates) # this is just a placeholder; we'll overwrite this as well, with summaries of the outcome tibble
  for(i in 1:replicates){
      outcome[,i] <- sample(c("heads", "tails"), size = trials, replace = T) # prob = c(0.55,0.45) <- make it not quite 50-50
      summary$all_tails[i] <- sum(outcome[,i] == "tails")
  }
  #print(outcome)
  ggplot(data = summary, aes(x = all_tails)) +                    # these lines are the function output
    geom_histogram(fill = "pink", binwidth = 0.5) +               # instead of printing something to the console
    xlim(0,trials) + theme_bw() +                                 # this time we'll have our function make us a graph!
    labs(x = str_c("number of tails in ", trials, " coin flips")) # custom label (fancy)
}


# The y-axis in the figures below represents the number of replicate trails that had some number of tails (x-axis) 
flip_coins(trials = 10, replicates = 10)
flip_coins(trials = 100, replicates = 100)
```

## VII. While loops

In general *while loops* come in handy less often but are still good to know about.

```{r while(){}}
# make a function that simulates flipping a coin once
flip <- function() sample(c("heads", "tails"), 1) # BTW you can write a function all on one line & skip the {}

# GOAL: Write a while loop that outputs the the number of flips it took to get three heads in a row

# First, set starting values for the variables we are using within our while() loop
nflips <- 0 # number of loops (coin flips) so far
nheads <- 0 # number of heads IN A ROW

while (nheads < 3) {       # when nheads == 3, the cycle will stop
  if (flip() == "heads") { # each time the "coin" lands heads up
    nheads <- nheads + 1   # add one to our running tally of heads
    } else {               # each time it doesn't (meaning tails up)
      nheads <- 0          # we have to start our nheads count over :(
      }
  nflips <- nflips + 1     # every time we go through a loop (one coin flip), add one
}


nflips 
#WHAT IS nflips GIVING US??
```

## VIII. Example exercise meant to get us thinking about some of the concepts we talked about in lecture

Whats the probability of rolling a "20" on a 20-sided die five times out of 10 rolls?

```{r example exercise, warning = FALSE}
k <- 5    # number of events
N <- 10   # number of trials
p <- 1/20 # probability of an event, == 0.05
dbinom(x = k, size = N, prob = p) # P(data|model) --dbinom gives us the absolute probability of getting 5 out of 10 rolls when P = 0.05 is really unlikely! 

# the code below gives us the probabilities of all possible outcomes for k (0 through 10) when P(20) == 1/20
probability <- tibble(ks = 0:N,
                      ps = dbinom(x = ks, size = N, prob = p)) # distribution of all possible outcomes 

### Theoretical distribution of probabilities
ggplot(probability, aes(x = ks, y = ps)) + # plots outcome on the x-axis and probability on the y-axis
  geom_col(fill = "red") + geom_vline(xintercept = k, linetype = "dashed") +   # you don't have to know
  labs(title = str_c("binomial distribution for trials = ", N,", prob = ", p), # the rest of this code
       x = "number of events", y = "probability") +                            # but feel free to dissect
  scale_x_continuous(breaks = c(0,2,4,6,8,10)) + theme_bw()                    # if you are curious
```


```{r example exercise, warning = FALSE}
# Compare dbinom() output to "experimentally" testing what we get on average when we roll 10 d20s many many times & P(20) == 1/20
trials <- 10                              
replicates <- 1000                        

outcome <- tibble(roll = 1:trials)        # placeholder -- we'll overwrite this, but need something there to overwrite 
summary <- tibble(all_20s = 1:replicates) # placeholder -- we'll overwrite this as well, with summaries of the outcome tibble

for(i in 1:replicates){
  outcome[,i] <- roll_d20(trials) # each cell of the outcome tibble is a d20 roll (function from above)
  summary$all_20s[i] <- sum(outcome[,i] == 20) # each cell here is the number of 20s rolled in one replicate
}

### Actual distribution of experimental outcomes
ggplot(data = summary, aes(x = all_20s)) +     # plots outcome on x-axis & makes histogram of the results                  
  geom_histogram(fill = "red", binwidth = 0.5) +               
  labs(title = str_c("actual outcome when trials = ", trials, " and replicates = ", replicates),
       x = "number of events", y = "number of replicates") +
  scale_x_continuous(breaks = c(0,2,4,6,8,10), limits = c(-0.45,10.45)) +
  geom_vline(xintercept = k, linetype = "dashed") + # events = 5 out of trials = 10
  theme_bw()
```


```{r example exercise, warning = FALSE}
# But let's say I *DID* roll five 20s out of 10 total rolls
# Maybe I have a magic d20 with a P(20) > 1/20...?
# BASED ON MY RESULTS (one dataset), what is the most likely model for P(20) using my die?

# The code below lets us answer that question by obtaining the likelihood of different probabilities of rolling a 20
likelihood <- tibble(models = seq(from = 0.01, to = 0.99, by = 0.01), # different probabilities in question
                     Lik_byhand = models^k*(1-models)^(N-k)*choose(N,k), # P(data|model) calculated by hand
                     Lik_dbinom = dbinom(x = k, size = N, prob = models))  # same thing obtained w/ dbinom()


ggplot(likelihood, aes(x = models, y = Lik_byhand)) + # switch out y = Lik_dbinom and see if it changes
  geom_line(color = "red") + geom_vline(xintercept = p, linetype = "dashed") + # line where P(20) = 1/20
  labs(title = str_c("most likely P(20) when trials = ", N, " and events = ", k),
       x = "model for P(20)", y = "P(model|data)") + theme_bw()
```


## IX. LAB EXERCISE

We want to know if the number of seedlings in year 0 (`seedlings.0`) is a function of seeds in year -1, -2, -3, or -4. 

In other words, we want to use our data to:
(1) Determine the most likely model for the number of seeds we counted this year based on << the number of seeds counted last year (year -1) >> for each of some number of hypothesized probability values. Recall that by the most "likely model" we mean the parameter value (here probability) that has the maximum likelihood
(2) Repeat step 1 above with the data for << the number of seeds we counted year before last (year -2) >> and << year -3 >> and << year -4 >>


Important information:

- The dbinom() function calculates the probability of getting some number of events out of some number of trials, given some probability of an event occurring in the first place. 
dbinom(x, size, prob)
x = k = number of events
size = N = number of trials
prob = p = probability of an event occurring

- Seedlings in year 0 is our `x` (*k*), or number of events.  
- Our size, *N*, is how many seeds we started out with in the previous years, depending on the lag (-1, -2, -3, or -4 years, i.e., the numbers of trials differs unless the starting number of seeds was the same in different years). 
- We want to calculate the probability of our observed data using a binomial distribution (given a particular model). 

Here our "model" has two parts: 

1. Our hypothesis about the lag, i.e., how many years (influences N)
2. Our binomial probability of a seed resulting in a seedling (p)

```{r lab exercise, read in data}
# read in data
orchids <- read_csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2024/Labs/Week3_functionsLoops/orchid.seedlings.csv")

#look at the data!!
orchids
```

```{r lab exercise, starting small}
# let's start with just one estimate for the probability of our data for just one of the hypothesized lags
# let's pretend (just guessing here) that the lag is four years and p = 0.05
#dbinom(no. events, no. trials, prob)
logp_4y <- dbinom(x = orchids$seedlings.0, 
                  size = orchids$seeds.m4, # seeds planted four years ago
                  prob = 0.05, log = TRUE) # last arguments means we'll get LOG likelihoods

logp_4y # vector of probabilities for each observation (rows in the data frame), given our model p = 0.05

logl_4y <- sum(logp_4y) # log-likelihood of the whole data set == the sum of the log-likelihoods of observations; remember we add instead of multiple because we're working with logs -- this is just a number that we can compare to other models

# SO... the log probability ~ log likelihood of getting our data for the number of seedlings in year 0 based on the number of seeds 4 years ago and a probability of germination of 0.5 ==
logl_4y
```


```{r lab exercise, repeat "starting small" for other years}
#Note, the above is just (1) the log-likelihood for ONE value of the probability of a seedling becoming a seed, and (2) for just ONE of our 4 hypotheses (-1, -2, -3 or -4 year lag)

# Now compare that to a model in which we think that the lag is three years instead (p is still 0.05)
logp_3y = dbinom(x = orchids$seedlings.0, 
                  size = orchids$seeds.m3, # seeds planted four years ago
                  prob = 0.05, log = TRUE)

# log probability ~ log likelihood of getting our data for the number of seedlings in year 0 based on the number of seeds 4 years ago and a probability of germination of 0.5 ==
(logl_3y = sum(logp_3y))
```

```{r lab exercise, compare lags}
logl_4y # log-likelihood of the whole data set given -4 year lag and p = 0.05
logl_3y # log-likelihood of the whole data set given -3 year lab and p = 0.05

# which model is more likely, do you think?
logl_4y/logl_3y
```

That was fun, but it could get a little repetitive...
INSTEAD, write code to calculate the log-likelihood of the whole data set given a range of 1) lags (-4, -3, -2, -1) and 2) probabilities of seeds becoming seedlings! (values between 0 - 1)
```{r lab exercise, complete your loop}

# I've made an empty tibble for you to start with, you just need to fill it in
# THEN try plotting the output given the code provided below

orchids_pt2 <- tibble(prob = seq(0.01, 0.99, by = 0.01),
                      lag_1y = 0,
                      lag_2y = 0,
                      lag_3y = 0,
                      lag_4y = 0)
# your
# code
# goes
# here

orchids_tidy <- orchids_pt2 %>% 
  pivot_longer(c(lag_1y,lag_2y,lag_3y,lag_4y), # remember this function?
               names_to = "population",  # it's so useful! :) :) :)
               values_to = "log.likelihood")

# Now PLOT your data also
ggplot(orchids_tidy, aes(x = probs, y = log.likelihood, color = population)) + 
  geom_line() + labs(x = "probability", y = "log-likelihood") + theme_bw() 

ggplot(orchids_tidy, aes(x = probs, y = exp(log.likelihood), color = population)) + 
  geom_line() + labs(x = "probability", y = "likelihood") + theme_bw()
```

SUBMIT THIS:
```{r submit this}
max(orchids_pt2$lag_1y) # log-likelihood at MLE, as calculated w/ a loop
index <- which.max(orchids_pt2$lag_1y) # row number in which you find MLE
orchids_pt2$prob[index] # MLE, i.e. probability that maximizes log-likelihood
```


The MLE and probabilities you came out with above should hopefully match the glm() output

YOU DON'T KNOW WHAT ANY OF THIS CODE BELOW MEANS AT THIS POINT. But, this code should allow you to CHECK your results above :)
```{r sanity check}
orchids$response <- cbind(orchids$seedlings.0, orchids$seeds.m1 - orchids$seedlings.0)
m1 <- glm(response ~ 1, family = binomial(link = "identity"), data = orchids) 

logLik(m1) # log-likelihood evaluated at MLE
coef(m1) # MLE of the coefficient (i.e. probability)

# Why don't they match perfectly? (hint: add '+ ylim(-500,0)' to the end of the log-likelihood ggplot above)
```


## Bonus Content:

If you don't have a good example data set to read in with read_csv(), and are tired of coming up with numbers haphazardly, you can generate random numbers from a given distribution instead

```{r BONUS CONTENT on generating data}
runif(n = 10, min = 0, max = 1) # same as runif(10) -- generates numbers from a *uniform* distribution
rnorm(n = 10, mean = 0, sd = 1) # tab is v. useful! -- generates numbers from a *normal* distribution


### don't worry about how to make this plot... YET! ###
fun_with_R <- tibble(sequence = 1:10000, 
                     uniform = runif(length(sequence), min = -4, max = 4), 
                     normal = rnorm(length(sequence)))
ggplot(fun_with_R) + 
  geom_histogram(aes(x = normal), alpha = 0.5, fill = "yellowgreen", bins = 100) + 
  geom_histogram(aes(x = uniform), alpha = 0.5, fill = "pink", bins = 100) +
  labs(x = "", y = "") + theme_bw()
### end of things you don't have to worry about ###


rbinom(n = 5, size = 10, prob = 0.5) # generates numbers from a *binomial* distribution
num_trials <- 100                    # now behold the power of variable assignment...
num_replicates <- 2000               # just change these & run the code below -- easy!


### start of things you don't have to worry about ###
fun_with_rbinom <- tibble(sequence = 1:(num_replicates*5),  
                          probability = rep(c("miniscule", "small", "fifty-fifty", "large", "omg"), 
                                            each = num_replicates),
                          binomial = c(rbinom(num_replicates, size = num_trials, prob = 0.01),
                                       rbinom(num_replicates, size = num_trials, prob = 0.1),
                                       rbinom(num_replicates, size = num_trials, prob = 0.5),
                                       rbinom(num_replicates, size = num_trials, prob = 0.9),
                                       rbinom(num_replicates, size = num_trials, prob = 0.99)))
ggplot(fun_with_rbinom, 
       aes(x = binomial, 
           fill = probability %>% factor(levels = c("miniscule", "small", "fifty-fifty", "large", "omg")))) +
  geom_histogram(binwidth = 1, position = "dodge") + scale_fill_discrete(name = "Probability of event") + 
  labs(x = "", y = "") + theme_bw() 
### end of things you don't have to worry about ###
```