---
title: "Somerville Gardens"
author: "Lee Brown"
date: "2025-09-20"
output: html_document
---

# Load libraries
```{r}
library(tidyverse)
library(car)
library(bbmle)
library(MASS)
```

# Drawing a pretty chi-square distribution
```{r chi-square distribution}
xvals <- seq(0,10, 0.1)

yvals.pdf <- dchisq(xvals, 1) # probability distribution
yvals.cdf <- pchisq(xvals, 1, lower.tail = T) # cumulative probability distribution

ggplot() +
  geom_line(aes(x = xvals,y = yvals.pdf), col = "red", size = 1) +
  theme_classic(base_size = 16) +
  geom_vline(xintercept = 2.8, linetype = "dashed", size = 1) +
  labs(x = "2 x ln(likelihood ratio)", y = "proportion of studies") 

ggplot() +
  geom_line(aes(x = xvals,y = yvals.cdf), col = "red", size = 1) +
  theme_classic(base_size = 16) +
  geom_vline(xintercept = 2.80, linetype = "dashed", size = 0.75) +
  geom_hline(yintercept = 0.95, col = "gray", size = 0.75) +
  geom_hline(yintercept = 1-0.094, linetype = "dashed", size = 0.75) +
  labs(x = "2 x ln(likelihood ratio)", y = "probability of X or lower") 

pchisq(2.80, df=1, lower.tail = F)
```

# Load and tidy garden data
```{r garden data}
(dat <- read.csv("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/LectureRcode/Garden_flowers.csv") %>% pivot_longer(cols = 2:17, names_to = "FlowerType"))

(dat1 <- dat %>% group_by(FlowerType, YardType) %>% summarize(total= sum(value)) %>% pivot_wider(names_from = "YardType", values_from = "total"))

dat2 <- dat1 %>% pivot_longer(c("Conventional","Pollinator"), names_to = "YardType", values_to = "yards.with")

dat2$yards.tot <- rep(50,nrow(dat2))

dat2 # check that dataframe looks the way we want it to
```

# define pollinator vs non-pollinator flowers
```{r pollinator vs nonpollinator plants}
#all flowers
(flowers <- unique(dat1$FlowerType))

#pollinator plant category
(poll.flwr <- c("Achillea","Asclepias","Aster","Campanula","Echinacea","Helianthus","Rudbeckia","Salvia"))

#non-pollinator plant category
(non.poll <- flowers[!flowers %in% poll.flwr])
```

# Study design
50 conventional gardens surveyed
50 pollinator gardens surveyed
All gardens classified via Google Streetview

Eight plants used to identify pollinator-friendly yards via Google Streetview

# Question
Does the proportion of yards with these pollinator-friendly plants differ between those classified as “pollinator-friendly” versus “conventional”?

Expected results: Yes, there should be a significantly higher proportion of the pollinator-friendly plants in yards classifed as "pollinator-friendly"

# Approach
First, do the statistics "by hand" (not using the glm() function) both for practice and because the glm crashes when a category is only composed of 0's. We will learn ways to correct this later, but for simple models, one can take this "old-school" method

# Single pollinator plant LLs
```{r aster LLs by hand}
# first test run for a single species, then create a loop to run for all
dat2
(asterdat <- dat2[dat2$FlowerType == "Aster",])

#single probability
(p.both = with(asterdat, sum(yards.with)/sum(yards.tot)))

# different probability for conventional vs pollinator-friendly yards
(p.sep = with(asterdat, yards.with/yards.tot))

# Get LLs for single model vs separate probability of seeing asters in pollinator-friendly vs conventional yards
(LL1 = sum(with(asterdat, dbinom(yards.with, yards.tot, p.both, log = T))))

(LL2 = sum(with(asterdat, dbinom(yards.with, yards.tot, p.sep, log = T))))

# Get statistic for comparing to a chi-square distribution
(chistat = 2*(LL2-LL1))

# Get p-value for comparing model with different probabilities for finding aster in conventional and pollinator-friendly yards vs single probability of seeing aster across all yard types
(pval = pchisq(q = chistat, df = 1, lower.tail = F))
```

# Check above value with GLM
```{r aster GLM}
# single probability of seeing aster
m1 <- glm(cbind(yards.with, 50-yards.with) ~ 1, family = binomial(link = "identity"), data = asterdat)

coef(m1) # compare to probabilities obtained in above code chunk (p.both vs p.sep)
logLik(m1) # compare to likelihoods calculated in above code chunk (LL1 vs LL2)

m2 <- glm(cbind(yards.with, 50-yards.with) ~ 0 + YardType, family = binomial(link = "identity"), data = asterdat)

coef(m2) # compare to probabilities obtained in above code chunk (p.both vs p.sep)
logLik(m2) # compare to likelihoods calculated in above code chunk (LL1 vs LL2)
```

# Calculate CIs from LL profile
```{r aster LL profile CIs}
myprobs <- seq(0,1, 0.001)

# CIs FOR CONVENTIONAL YARDS
# first get LLs for all possible probabilities
my.profile.conv <- with(asterdat[1,], dbinom(yards.with, yards.tot, myprobs, log = T))

# Get CIs for probability of seeing asters in conventional yards
# Recall that all values within 1.92 of the max LL value are not significantly different from the MLE and so fall within the CIs
(CI.conv <- myprobs[which(my.profile.conv > (max(my.profile.conv)-1.92))])
  
# CIs FOR POLLINATOR-FRIENDLY YARDS
my.profile.poll <- with(asterdat[2,], dbinom(yards.with, yards.tot,myprobs, log = T))

# Get CIs for probability of seeing asters in pollinator-friendly yards
(CI.poll <- myprobs[which(my.profile.poll > (max(my.profile.poll)-1.92))])
```

# Plot means and CIs for aster
```{r plot aster probabilities}
# Create a dataframe for plotting using ggplot
aster_ci <- data.frame(
  group = asterdat$YardType,
  mean_val = p.sep,
  lower_ci = c(min(CI.conv),min(CI.poll)),
  upper_ci = c(max(CI.conv),max(CI.poll))
)

ggplot(aster_ci, aes(x = group, y = mean_val)) +
  geom_point(color = "darkorchid", size = 3) + # Plot the mean values
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.1, color = "darkorchid", size = 0.8) + # Add error bars
  labs(x = "Yard type", y = "Proportion of yards") +
  annotate("text", x = 0.7, y = 0.45, label = expression(chi^2)) +
  annotate("text", x = 0.83, y = 0.44, label = " = 2.80") +
  annotate("text", x = 0.8, y = 0.41, label = "P = 0.09")  +
  theme_classic(base_size = 18)
```

# Plot cumulative probabilites for aster to visualize CIs
```{r plot all results}
# Cumulative probabilities for pollinator-friendly yards
ggplot() +
  geom_line(aes(x = myprobs[2:1000],y = my.profile.poll[2:1000]), size = 1) + #[2:1000] to remove the -Inf values for the LL at 0 and 1
  theme_classic(base_size = 16) +
  geom_vline(xintercept = min(CI.poll), linetype = "dashed", color = "blue",size = 0.75) +
  geom_vline(xintercept = max(CI.poll), linetype = "dashed", color = "blue", size = 0.75) +
  geom_hline(yintercept = max(my.profile.poll)-1.92, col = "green4", size = 1) +
  labs(x = "possible proportion of\n pollinator-friendly yards with asters", y = "ln[L(model|data)]") 

# Cumulative probabilities for conventional yards
ggplot() +
  geom_line(aes(x = myprobs[2:1000],y = my.profile.conv[2:1000]), col = "darkgray",size = 1) + #[2:1000] to remove the -Inf values for the LL at 0 and 1
  theme_classic(base_size = 16) +
  geom_vline(xintercept = min(CI.conv), linetype = "dashed", color = "blue",size = 0.75) +
  geom_vline(xintercept = max(CI.conv), linetype = "dashed", color = "blue", size = 0.75) +
  geom_hline(yintercept = max(my.profile.conv)-1.92, col = "green4", size = 1) +
  labs(x = "possible proportion of\n conventional yards with asters", y = "ln[L(model|data)]") 
```

# Loop over all pollinator-friendly yards to calculate means/CIs and p-values and plot results to file
```{r calculate LLs looped over flower types}
AllFlowers <- c(poll.flwr,non.poll)

color.flowers <- c("darkolivegreen3", "darkorange","darkorchid", "cornflowerblue", "orchid", "goldenrod", "chocolate4", "midnightblue", "lightgoldenrod","gold","skyblue", "magenta","pink3","tomato","red", "purple")


myprobs <- seq(0,1, 0.001)

mystats <- data.frame(
  plant = AllFlowers,
  chixchi = NA,
  pvalue = NA,
  mean.poll = NA,
  mean.conv = NA,
  lowerCI.poll = NA,
  upperCI.poll = NA,
  lowerCI.conv = NA,
  upperCI.conv = NA)

plotresults <- data.frame(
  group = tempdat$YardType,
  mean_val = NA,
  lower_ci = NA,
  upper_ci = NA)

for(i in 1:length(AllFlowers)){
# temp df to calculate values
  tempdat <- dat2[dat2$FlowerType == AllFlowers[i],] 
# calculate statistics
  p.both <- with(tempdat, sum(yards.with)/sum(yards.tot))
  p.sep <- with(tempdat, yards.with/yards.tot)
  LL1 <- sum(with(tempdat, dbinom(yards.with, yards.tot, p.both, log = T)))
  LL2 <- sum(with(tempdat, dbinom(yards.with, yards.tot, p.sep, log = T)))
  chistat <- 2*(LL2-LL1)
  pval <- pchisq(q = chistat, df = 1, lower.tail = F)
# get LLs for all possible probabilities
  my.profile.conv <- with(tempdat[1,], dbinom(yards.with, yards.tot, myprobs, log = T))
  my.profile.poll <- with(tempdat[2,], dbinom(yards.with, yards.tot,myprobs, log = T))
# get CIs
  CI.conv <- myprobs[which(my.profile.conv > (max(my.profile.conv)-1.92))]
  CI.poll <- myprobs[which(my.profile.poll > (max(my.profile.poll)-1.92))]
# fill values in df:
  mystats$chixchi[i] <- chistat
  mystats$pvalue[i] <- pval
  mystats$mean.conv[i] <- p.sep[1]
  mystats$mean.poll[i] <- p.sep[2]
  mystats$lowerCI.poll[i] <- min(CI.poll)
  mystats$upperCI.poll[i] <- max(CI.poll)
  mystats$lowerCI.conv[i] <- min(CI.conv)
  mystats$upperCI.conv[i] <- max(CI.conv)
# fill in df for plotting results
  plotresults$mean_val <- p.sep
  plotresults$lower_ci <- c(min(CI.conv),min(CI.poll))
  plotresults$upper_ci <- c(max(CI.conv),max(CI.poll))
# plot results
 ggplot.results <-
    ggplot(plotresults, aes(x = group, y = mean_val)) +
    ylim(0,0.75) +
    geom_point(color = color.flowers[i], size = 3) + # Plot the mean values
    geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.1, color = color.flowers[i], size = 0.8) + # Add error bars
    labs(x = "Yard type", y = "Proportion of yards") +
    annotate("text", x = 0.7, y = 0.68, label = expression(chi^2)) +
    annotate("text", x = 0.83, y = 0.68, label = paste(" = ", round(mystats$chixchi[i],digits = 3), sep = "")) +
    annotate("text", x = 0.77, y = 0.65, label = paste("P = ", round(mystats$pvalue[i],digits = 3), sep = ""))  +
    theme_classic(base_size = 18)
  ggsave(paste("/Users/brown8lm/Dropbox/JMU/courses/BIO456_EcologicalModels&Data/Fall2025/Labs/Lab6/figures/",AllFlowers[i],".pdf",sep = ""),ggplot.results)
}
```

# Compare AIC and LRT - with shortcuts that we will use later
```{r AIC and LRT comparisons}
myout <- array(NA, dim = c(16,4))

for(i in 1:16){
  usedat <- dat[dat$FlowerType == AllFlowers[i],]
  w1 <- glm(value ~ 1, family = binomial, data = usedat) #usedat[complete.cases(usedat)==T,])
  w2 <- glm(value ~ YardType, family = binomial, data = usedat) #usedat[complete.cases(usedat)==T,])
  myout[i,] <- unlist(c(Anova(w2)[c(1,3)], ICtab(w1,w2, sort = F)[1]))
}

myout2 <- as_tibble(myout)
names(myout2) <- names(unlist(c(Anova(w2)[c(1,3)], ICtab(w1,w2, sort = F)[1])))
myout2$FlowerType <- AllFlowers

myout2
```

# Competing models for the probability of roses in yards
1) Same probability of roses in all yards
2) Probability of roses depends on yard type (conventional vs. pollinator-friendly)
3) Probability of roses depends on neighborhood
4) Probability of roses depends on latitude (north-south)
5) Probability of roses depends on longitude (east-west)

Model 1: Same probability of roses in all yards
```{r roses - competing models - single p}
# limit data to rows for which the flower type is a Rose
usedat <- dat[dat$FlowerType == "Rose",]

# Note that in this version of the data each row is a 0 or 1
# That's okay for glm(), we just don't need the cbind() function

z1 <- glm(value ~ 1, family = binomial(link = "identity"), data = usedat[complete.cases(usedat)==T,])

(z1mean <- coef(z1))
logLik(z1)

(z1confint <- confint(z1))
AIC(z1)

plotresults <- data.frame(
  group = "all sites",
  mean_val = as.numeric(coef(z1)),
  lower_ci = as.numeric(z1confint[1]),
  upper_ci = as.numeric(z1confint[2]))

ggplot(plotresults, aes(x = group, y = mean_val)) +
  geom_point(color = "tomato", size = 3) + 
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.1, color = "tomato", size = 0.8) +
  ylim(0,0.9) +
  labs(x = "Yard type", y = "Proportion of yards\n with roses") +
  theme_classic(base_size = 18)
```

Model 2: Probability differs by yard type
```{r roses - competing models - yard type}
z2 <- glm(value ~ 0 + YardType, family = binomial(link = "identity"), data =  usedat[complete.cases(usedat)==T,])

coef(z2)
logLik(z2)
confint(z2)
AIC(z2)

# Plot me on your own!
plotCI(1:2, coef(z2), li = confint(z2)[,1], ui = confint(z2)[,2], xlim = c(0.5,2.5), ylim = c(0,1), xaxt = "n", ylab = "", xlab = "", pch = 19, col = 'tomato', lwd = 3)
  axis(side = 1, at = 1:2, labels = c("Conv.", "Poll."))
  mtext(side = 1, line = 2, "Yard Type")
  mtext(side = 2, line = 2, "prop. yards with roses")
```

Model 3: Probability differs by neighborhood
```{r roses - competing models - neighborhood}
z3 <- glm(value ~ 0+NEIGHBORHOOD, family = binomial(link = "identity"), start=rep(0.01,12), data =  usedat[complete.cases(usedat)==T,])

logLik(z3)
AIC(z3)

  
# Shenanigans to get plot of values for each group
myprobs <- seq(0,1, 0.001)
mylevels <- unique(usedat$NEIGHBORHOOD)[order(unique(usedat$NEIGHBORHOOD))]  
my.prof <- array()
low.CI <- array()
upp.CI <- array()
MLEs <- array()

for(i in 1:length(mylevels)){
  usedat2 <- dat[dat$NEIGHBORHOOD == mylevels[i],]
  for(j in 1:length(myprobs)){
    my.prof[j] <- with(usedat2, sum(dbinom(value, 1, myprobs[j], log = T)))
  }
  my.CI = myprobs[which(my.prof > (max(my.prof)-1.92))]
  MLEs[i] = myprobs[which(my.prof == max(my.prof))]
  low.CI[i] = min(my.CI)
  upp.CI[i] = max(my.CI)
}

low.CI
upp.CI
MLEs
# compare MLEs to:
coef(z3)

# Convert the plot below to ggplot code
plotCI(1:12, MLEs, li = low.CI, ui = upp.CI, xlim = c(0.5,12.5), ylim = c(0,1), xaxt = "n", ylab = "", xlab = "", pch = 19, col = 'tomato', lwd = 3)
## Draw the x-axis labels.
text(x = 1:length(mylevels),
     ## Move labels to just below bottom of chart.
     y = par("usr")[3] - 0.02,
     ## Use names from the data list.
     labels = mylevels,
     ## Change the clipping region.
     xpd = NA,
     ## Rotate the labels by 35 degrees.
     srt = 25,
     ## Adjust the labels to almost 100% right-justified.
     adj = 0.965, cex = 0.9)
mtext(side = 2, line = 2, "prop. yards with roses")
```

Model 4: Probability is a function of latitude
```{r roses - competing models - latitude}
z4 <- glm(value ~ lat, family = binomial(link = "identity"), data = usedat[complete.cases(usedat)==T,])

coef(z4)
logLik(z4)
AIC(z4)
  
# model 4 - latitude
minLat <- min(usedat$lat, na.rm = T)
maxLat <- max(usedat$lat, na.rm = T)
mylats <- seq(minLat, maxLat, length.out=100)
mypreds <- predict(z4, newdata = data.frame(lat = mylats), se.fit = T)

# Convert the plot below to ggplot code
plot(mylats, mypreds$fit, ylim = c(0, 1), xlab = "", ylab = "", type = "l", lwd = 4, col = "tomato")
#points(mylats, mypreds$fit - 1.96*mypreds$se.fit, col = "tomato", type = "l", lwd = 2, lty = "dotted")
#points(mylats, mypreds$fit + 1.96*mypreds$se.fit, col = "tomato", type = "l", lwd = 2, lty = "dotted")
points(usedat$lat, usedat$value)
mtext(side = 1, line = 2, "Latitude")
#mtext(side = 1, line = 3, at = c(minLat, maxLat), c("south", "north"))
mtext(side = 2, line = 2, "prop. yards with roses")
```

Model 5: Probability is a function of longitude
```{r roses - competing models - longitude}
z5 <- glm(value ~ long, family = binomial(link = "identity"), data =  usedat[complete.cases(usedat)==T,])

coef(z5)
logLik(z5)
AIC(z5)
  
# model 5 - longitude
minLon = min(usedat$long, na.rm = T)
maxLon = max(usedat$long, na.rm = T)
mylons = seq(minLon, maxLon, length.out=100)
mypreds = predict(z5, newdata = data.frame(long = mylons), se.fit = T)

# Convert the plot below to ggplot code
plot(mylons, mypreds$fit, ylim = c(0, 1), xlab = "", ylab = "", type = "l", lwd = 4, col = "tomato")
#points(mylons, mypreds$fit - 1.96*mypreds$se.fit, col = "tomato", type = "l", lwd = 2, lty = "dotted")
#points(mylons, mypreds$fit + 1.96*mypreds$se.fit, col = "tomato", type = "l", lwd = 2, lty = "dotted")
points(usedat$long, usedat$value)
mtext(side = 1, line = 2, "Longitude")
#mtext(side = 1, line = 3, at = c(minLon, maxLon), c("west", "east"))
mtext(side = 2, line = 2, "prop. yards with roses")
```

Exploring LL's, AIC's & Akaike weights
```{r}
# LLs for each model
(z_LLs <- round(rbind(logLik(z1),logLik(z2),logLik(z3),logLik(z4),logLik(z5)),2))

# Number of parameters for each model
(Nparam_LLs <- c(length(coef(z1)),length(coef(z2)),length(coef(z3)),length(coef(z4)),length(coef(z5))))

# Convert LLs to AICs (as proof of concept)
2 * Nparam_LLs - 2 * z_LLs

# Get AIC from model using AIC() function
# Display comparison of model AICs using ICtab() function

# Different ways of showing output from ICtab
ICtab(z1,z2,z3,z4,z5, delta = F, base = T, sort = F)
ICtab(z1,z2,z3,z4,z5, sort = F)
ICtab(z1,z2,z3,z4,z5, sort = F, weights = T)
ICtab(z1,z2,z3,z4,z5, base = T, weights = T)

# Numerator for each model's Akaike weight
exp(-0.5*ICtab(z1,z2,z3,z4,z5, sort = F)$dAIC)
# Sum of the numerators
sum(exp(-0.5*ICtab(z1,z2,z3,z4,z5, sort = F)$dAIC))



#Comparing models 3 & 1
2*(logLik(z3) - logLik(z1))
```

Comparing models 3 & 1 with LRT
```{r comparing models 3 & 1 - PDF & CDF}

# chi-square distribution with 11 df
xvals = seq(0,40, 0.1)
yvals.pdf = dchisq(xvals, 11) # probability distribution (PDF)
yvals.cdf = pchisq(xvals, 11, lower.tail = T) # cumulative probability distribution (CDF)

# highlight all 4 lines of code below at once to show plot
plot(xvals, yvals.pdf, xlab = "", ylab = "", type = "l", col = "red", lwd = 3)
mtext(side = 1, line = 2, "2 x ln(likelihood ratio)")
mtext(side = 2, line = 2, "prop. of studies")
abline(v = 27.3, col = "blue")

# highlight all 5 lines of code below at once to show plot
plot(xvals, yvals.cdf, xlab = "", ylab = "", type = "l", col = "red", lwd = 3, ylim = c(0,1))
mtext(side = 1, line = 2, "2 x ln(likelihood ratio)")
mtext(side = 2, line = 2, "prob. of X or lower")
abline(h = 0.95, col = "gray")
abline(v = 27.3, col = "blue")


pchisq(27.3, df=11, lower.tail = F)
```

