---
title: "Lecture 2 CHALLENGE #2"
author: "Lee Brown"
date: "2025-08-25"
output: html_document
---

```{r}
library(ggplot2)
```

1. Calculate the likelihood and log-likelihood of survival having each of the following values: 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9 using the 5-plant data set

2. Plot the likelihood (y-axis) vs. value of survival (x-axis)

3. Repeat your calculations and plotting using the 10-plant data set.  

Create a vector of survival alues
```{r survival values}
#probability and likelihood lecture
surv<-c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9)
#or
surv<-seq(0.1,0.9,0.1)

surv
```

Calculate likelihoods
```{r likelihoods}
#calculate likelihoods
fiveplants<-surv*surv*surv*surv*(1-surv) #or surv^4*(1-surv)
tenplants<-surv^8*(1-surv)^2

#check that the calculated values make sense
fiveplants 
tenplants 

#calculate log-likelihoods
fiveplantslog<-4*log(surv)+log(1-surv)
tenplantslog<-8*log(surv)+2*log(1-surv)

#check that the calculated values make sense
fiveplantslog
tenplantslog

#Note that there is zero likelihood that survival is either 0 or 1 because we already know at least 1 survives and at least 1 dies for both of our data sets.

#ALSO:
log(surv)
log(0)
log(1) #0
```

Make a dataframe for plotting the data in ggplot
```{r dataframe for plotting}
plantdat <- data.frame(probs = surv, dataset = rep(c(rep("fiveplants", length(surv)), rep("tenplants", length(surv))),2), likelihood = c(rep("like",18),rep("logL", 18)), vals = c(fiveplants, tenplants, fiveplantslog,tenplantslog))

plantdat #check dataframe
```

Plot the likelihood and log-likelihood
```{r plot the data}
#Plot likelihoods
ggplot(plantdat[plantdat$likelihood=="like",], aes(x = probs, y = vals, color = dataset)) +
         geom_line() +
  ylim(0,0.1) +
  labs(x = "survival value", y = "likelihood of dataset") +
  theme_classic(base_size = 18)

#Plot log-likelihoods
ggplot(plantdat[plantdat$likelihood=="logL",], aes(x = probs, y = vals, color = dataset)) +
         geom_line() +
  ylim(-20,0) +
  labs(x = "survival value", y = "likelihood of dataset") +
  theme_classic(base_size = 18)
```

A much simpler way of platting!! 
No need to create a dataframe, just plot the original vector data
```{r simpler plot with less coding}
#plot likelihoods
#highlight all lines of code below at the same time or add a "+" after each line
plot(surv,fiveplants,type="l", col = "blue", ylim=c(0,0.1),xlab="survival value", ylab = "likelihood of dataset")
lines(surv,tenplants,type="l",col="red")
legend("topleft", legend = c("five plants", "ten plants"), col = c("red", "blue"), bty = "n", lty = c(1, 1))

#plot log-likelihoods
#highlight all lines of code below at the same time or add a "+" after each line
plot(surv, fiveplantslog, type="l", col = "blue", ylim=c(-20,0), xlab = "survival", ylab = "log-likelihood")
lines(surv,tenplantslog,type="l",col="red")
legend("topleft", legend = c("five plants", "ten plants"), col = c("red", "blue"), bty = "n", lty = c(1, 1))

```


